Controlled Experiments for Word Embeddings

### 2015-09-03 article/main.tex
moderate adriaan changes; remove tf; varied plot markers

\begin{section}{Introduction}
The field of natural language processing (NLP) is presently undergoing rapid
developments, not in the least triggered by the tremendous commercial
promises it holds.  A timely research topic in NLP are word embeddings,
where words from a vocabulary are represented by dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e., given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

@@

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new tokens into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modelling the frequency and co-occurrence distributions of these tokens
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of tokens that all appear in the same context, but with different
frequencies, or study a family of tokens that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by inspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

@@

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}, who also based their work on word co-occurrences.
They define contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence $\mathrm{D} \big(\p(w') ||\p(w'|w) \big)$
between the distribution of context words, $\p(w')$, and the
co-occurrence distributions of that target word $w$.  Specifically,
\begin{equation} 
\label{cd}
\mathrm{D} \big(\p(w') ||\p(w'|w) \big) = \sum_{w'} \p(w'|w) \log_2
\frac{\p(w'|w)}{\p(w')},
\end{equation} 
where the sum is over the set of \textcolor{blue}{content?} words $w'$.  They set the
window size to five words before and five words after the target word.
To reduce the computational burden, they recorded co-occurrences of 690
target words, randomly chosen from equally divided frequency bands, with
only the $n=500$ most frequent content words.  As pointed out by the
authors, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a
word $w$ tends to appear in a wide variety of contexts, the distribution
of the words it occurs with, i.e., the posterior distribution $\p(w'|w)$,
tends to diverge less from the prior distribution $\p(w')$, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, the
posterior distribution tends to diverge more from the prior, and a
larger value of contextual distinctiveness results.  As for
word significance, contextual distinctiveness summarizes a word's
contextual behavior in a single number. \textcolor{blue}{shorten?}

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
They built a vocabulary from the 8k most frequent nouns and 4k most
frequent adjectives in a large general language corpus and added 22k
adjective-noun composites.  For each item in the vocabulary, they
recorded their co-occurrence within the same sentence with any of the
top 10k most frequent content words (nouns, adjectives or verbs) in the
corpus, and contructed word embeddings (of dimension 300) through
singular value decomposition of the co-occurrence matrix.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two
simplest being adding and component-wise multipying the adjective and
noun vectors.  They hypothesized that the length of the vectors thus
constructed can be used to distinguish acceptable and semantically
deviant adjective-noun composites.  Using a few hundred adjective-noun
composites selected by humans for evaluation, they found that deviant
composites have a shorter vector than acceptable ones, in accordance
with their expectation.

In the word2vec similarity and word relationship tasks, normalised word vectors are used.
Word vector length is thus disregarded.
In the word2vec forum\footnote{\googlegroup}, it has been observed that normalisation improves performance on these tasks since word vector length is related to word frequency.
It remained unclear, however, whether word frequency was \textit{directly} related to word vector length or rather via correlated factors such as word significance.
Our experimental approach demonstrates that a direct relationship indeed exists.
It is shown moreover that the normalisation of the word vectors also prevents word co-occurrence noise from affecting the performance of the word similarity and word relationship tasks.

On the other hand, recent theoretical work \cite{Arora2015} has approached the problem of explaining the so-called ``compositionality'' property exhibited by some word embeddings.
In that work, unnormalised vectors are used in their model of the word relationship task.
It is hoped that experimental approaches such as those described here might enable theoretical investigations to describe the role of the word vector length in the word relationship tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia datadump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word- (or ``term-'') frequency $tf$ is described in overview
in Table~\ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

\begin{subsection}{Word2vec}\label{word2vec}
Word2vec, a feedforward neural network with a single hidden
layer, learns word vectors from word co-occurrences in an unsupervised
manner.  Word2vec comes in two versions.  In the continuous bag-of-words
(CBOW) model, the words appearing around a target word serve as input.
That input is projected linearly onto the hidden layer and the networks
then attempts to predict the target word on output.  Training is
achieved through backpropagation.  The word vectors are encoded in the
weights of the first synaptic layer, ``syn0''.  The weights of the
second synaptic layer (``syn1neg'', in the case of negative sampling) are typically discarded.  In the other model,
called skip-gram, target and context words swap places, so that the
target word now serves as input, while the network attempts to predict
the context words on output.

@@

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
%
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table~\ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section ~\ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
number of occurrences sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum occurrence count of any token in
the corpus.  Figure~\ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Term frequency $tf$ of words chosen for the word
          frequency experiment. }
	\label{word-frequency-counts}
\end{table}

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section ~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
%
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we insperse the word uniformly at random throughout the corpus so
that its global frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the global word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher global frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of iserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure ~\ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a global frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

\subsection{Experimental results}
We next present the results of the word frequency experiment. We
consider the effect of word frequency on the direction and on the length
of word vectors separately.

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
%
Figure~\ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remain above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined number of
occurences equals the term frequency of that word in the original
corpus.  The red blocks in Figure ~\ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure ~\ref{word-frequency-experiment-heatmap}, where it is apparent that
the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these tokens.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.  In the
discussions below, the tokens \word{VOID\_i} and
\word{THE\_i} whose vectors have a cosine similarity less than $0.8$
with \word{VOID\_1} and \word{THE\_1}, respectively are therefore
discarded.

From these results, we conclude that word frequency has no effect on the
direction of a word vector, given a sufficient number of samples to
learn the co-occurrence distribution of the word.
%For well-learned representations, word frequency has no effect on word
%vector direction.

\subsubsection{Word frequency and vector length}
%
We next consider the effect of frequency on word vector length.
Figure~\ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the number of occurences of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure ~\ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e., the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure ~\ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $tf=10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant.  Since we already found the
direction to be also constant, it is sensible to speak of the word
vector of \word{VOID} irrespective of its frequency.  In particular, the
token \word{VOID\_1} may be taken as an approximation.

@@

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          tokens \word{THE\_i}, and their cosine similarity with the
          most frequent such token, \word{THE\_1}.  It is apparent from
          the nearest neigbour list that the vectors of the
          low-frequency tokens have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs. frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section ~\ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are
          included for completeness.  }
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Wheras in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurence, we next study the effect of co-occurence for
fixed frequency.

\subsection{Generating noise}
We do so again in a controlled manner by adding noise to the
co-occurence distribution.  We take the noise distribution to be the
global word distribution, so that noise can be simply added to the
co-occurrence distribution of a word by interspersing occurrences of
that word uniformly at random throughout the corpus.  A word that is
consistently used in a distinctive context in the unmodified corpus thus
appears in the modified corpus also in completely unrelated contexts.
As in Section \ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.  Table
\ref{cooccurrence-noise-words} lists the words chosen, along with their
frequencies in the corpus.

@@

The proportion of occurrences of the token \word{CAT\_i} that arise by
replacing the word \word{cat} is $P_{5/6, 8}(i)$, and the proportion
that arise from randomly interspersing this token throughout the corpus
is $r_i := 1 - P_{5/6, 8}(i)$.  The particular value of $p$ (large
compared to that of section \ref{WFVE}) was chosen so that the
proportion of noise $r_i$ did not increase too quickly with $i$.  These
proportions are the sampling points on our axis of variation, and this
value of $p$ (together with the choice of $n$) assures a reasonable
spread.  Other parameter values (or indeed other distributions) could,
of course, have been used equally well.

\begin{table}
	\input{noise-cooccurrence-experiment-counts.tex}
	\label{fig:co-occurrence-noise-counts}
	\caption{Term frequency $tf$ of words chosen for the
          co-occurrence noise experiment. }
	\label{cooccurrence-noise-words}
\end{table}

>>>>

\begin{section}{Introduction}
The field of natural language processing (NLP) is presently undergoing rapid
developments, not in the least triggered by the tremendous commercial
promises it holds.  A timely research topic in NLP are word embeddings,
where words from a vocabulary are represented by dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e. given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

@@

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new tokens into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modeling the frequency and co-occurrence distributions of these tokens
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of tokens that all appear in the same context, but with different
frequencies, or study a family of tokens that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by interspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

@@

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}, who also based their work on word co-occurrences.
They define contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence $\mathrm{D} \big(\p(w') ||\p(w'|w) \big)$
between the distribution of context words, $\p(w')$, and the
co-occurrence distributions of that target word $w$.  Specifically,
\begin{equation} 
\label{cd}
\mathrm{D} \big(\p(w') ||\p(w'|w) \big) = \sum_{w'} \p(w'|w) \log_2
\frac{\p(w'|w)}{\p(w')},
\end{equation} 
where the sum is over the set of \textcolor{blue}{content?} words $w'$.
As pointed out by the
authors, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a
word $w$ tends to appear in a wide variety of contexts, the distribution
of the words it occurs with, i.e. the posterior distribution $\p(w'|w)$,
tends to diverge less from the prior distribution $\p(w')$, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, the
posterior distribution tends to diverge more from the prior, and a
larger value of contextual distinctiveness results.

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
For each item in the vocabulary, they
recorded their co-occurrence within the same sentence with 
 frequent content words (nouns, adjectives or verbs) in the
corpus, and constructed word embeddings via 
singular value decomposition of the co-occurrence matrix.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two
simplest being adding and component-wise multiplying the adjective and
noun vectors.  They hypothesized that the length of the vectors thus
constructed can be used to distinguish acceptable and semantically
deviant adjective-noun composites.  Using a few hundred adjective-noun
composites selected by humans for evaluation, they found that deviant
composites have a shorter vector than acceptable ones, in accordance
with their expectation.

In the word2vec similarity and word relationship tasks, normalised word vectors are used.
Word vector length is thus disregarded.
In the word2vec forum\footnote{\googlegroup}, it has been observed that normalisation improves performance on these tasks since word vector length is related to word frequency.
It remained unclear, however, whether word frequency was \textit{directly} related to word vector length or rather via correlated factors such as word significance.
Our experimental approach demonstrates that a direct relationship indeed exists.
It is shown moreover that the normalisation of the word vectors also prevents word co-occurrence noise from affecting the performance of the word similarity and word relationship tasks.
On the other hand, recent theoretical work \cite{Arora2015} has approached the problem of explaining the so-called ``compositionality'' property exhibited by some word embeddings.
In that work, unnormalised vectors are used in their model of the word relationship task.
It is hoped that experimental approaches such as those described here might enable theoretical investigations to describe the role of the word vector length in the word relationship tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word occurrence counts are summarised 
in Table \ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

\begin{subsection}{Word2vec}\label{word2vec}
Word2vec, a feed-forward neural network with a single hidden
layer, learns word vectors from word co-occurrences in an unsupervised
manner.  Word2vec comes in two versions.  In the continuous bag-of-words
(CBOW) model, the words appearing around a target word serve as input.
That input is projected linearly onto the hidden layer and the networks
then attempts to predict the target word on output.  Training is
achieved through back-propagation.  The word vectors are encoded in the
weights of the first synaptic layer, ``syn0''.  The weights of the
second synaptic layer (``syn1neg'', in the case of negative sampling) are typically discarded.  In the other model,
called skip-gram, target and context words swap places, so that the
target word now serves as input, while the network attempts to predict
the context words on output.

@@

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table \ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section \ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
number of occurrences sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum occurrence count of any token in
the corpus.  Figure \ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Words chosen for the word frequency variation experiment, along with their occurrence counts in the unmodified corpus. }
	\label{word-frequency-counts}
\end{table}

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section \ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus so
that its global frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the global word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure \ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher global frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure \ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a global frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

\subsection{Experimental results}\label{WFVE-results}
We next present the results of the word frequency experiment. We
consider the effect of word frequency on the direction and on the length
of word vectors separately.

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
Figure \ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remain above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined number of
occurrences equals the term frequency of that word in the original
corpus.  The red blocks in Figure \ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure \ref{word-frequency-experiment-heatmap}, where it is apparent that
the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these tokens.
Table \ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.  In the
discussions below, the tokens \word{VOID\_i} and
\word{THE\_i} whose vectors have a cosine similarity less than $0.8$
with \word{VOID\_1} and \word{THE\_1}, respectively are therefore
discarded.

From these results, we conclude that word frequency has no effect on the
direction of a word vector, given a sufficient number of samples to
learn the co-occurrence distribution of the word.

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Figure \ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the number of occurrences of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e. the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure \ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant.  Since we already found the
direction to be also constant, it is sensible to speak of the word
vector of \word{VOID} irrespective of its frequency.  In particular, the
token \word{VOID\_1} may be taken as an approximation.

@@

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          tokens \word{THE\_i}, and their cosine similarity with the
          most frequent such token, \word{THE\_1}.  It is apparent from
          the nearest neighbour list that the vectors of the
          low-frequency tokens have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs. frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section \ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are
          included for completeness.  }
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Whereas in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurrence, we next study the effect of co-occurrence for
fixed frequency.

\subsection{Generating noise}
We do so again in a controlled manner by adding noise to the
co-occurrence distribution.  We take the noise distribution to be the
global word distribution, so that noise can be simply added to the
co-occurrence distribution of a word by interspersing occurrences of
that word uniformly at random throughout the corpus.  A word that is
consistently used in a distinctive context in the unmodified corpus thus
appears in the modified corpus also in completely unrelated contexts.
As in Section \ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.  Table
\ref{cooccurrence-noise-words} lists the words chosen, along with their
frequencies in the corpus.

@@

The proportion of occurrences of the token \word{CAT\_i} that arise by
replacing the word \word{cat} is $P_{5/6, 8}(i)$, and the proportion
that arise from randomly interspersing this token throughout the corpus
is $r_i := 1 - P_{5/6, 8}(i)$.  The particular value of $p$ (large
compared to that of Section \ref{WFVE}) was chosen so that the
proportion of noise $r_i$ did not increase too quickly with $i$.  These
proportions are the sampling points on our axis of variation, and this
value of $p$ (together with the choice of $n$) assures a reasonable
spread.  Other parameter values (or indeed other distributions) could,
of course, have been used equally well.

\begin{table}
	\input{noise-cooccurrence-experiment-counts.tex}
	\label{fig:co-occurrence-noise-counts}
	\caption{Words chosen for the co-occurrence noise experiment, along with the occurrence counts in the unmodified corpus. }
	\label{cooccurrence-noise-words}
\end{table}
